---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GenErlangFit <a href="https://vivekmurals.github.io/GenErlangFit"><img src="man/figures/logo.svg" align="right" height="139" alt="GenErlangFit Logo"></a>

<!-- badges: start -->
<!-- badges: end -->

GenErlangFit is an R package and web application for fitting Erlang and Erlang-Exponential distributions to empirical waiting time data, helping modelers incorporate realistic non-exponential waiting time distributions in compartmental models via the Linear Chain Trick.

The web application is available at https://vivek-murals.shinyapps.io/GenErlangFitApp/ for users who prefer an interactive interface.

## Installation

You can install the development version of GenErlangFit from [GitHub](https://github.com/vivekmurals/GenErlangFit) with:

``` r
install.packages("devtools") # If not already installed.
devtools::install_github("vivekmurals/GenErlangFit", build_vignettes = TRUE)
```

## Documentation

For detailed instructions on using the package, see the [Getting Started](https://vivekmurals.github.io/GenErlangFit/articles/GettingStarted.html) guide, which walks through the main function and its various arguments. This can also be accessed from an R session with:

```
vignette("GettingStarted", package = "GenErlangFit")
```
Technical details regarding the fitting algorithms and assumptions can be found in [Technical Details](https://vivekmurals.github.io/GenErlangFit/articles/TechnicalDetails.html).

## Data Format and Quick Start

The primary function `GenErlang_Fit` requires at minimum a single numeric vector of observed time-to-event values where all entries must be positive. In this documentation, we refer to this vector as `empiricaldata`.

Once your data is ready, fitting an Erlang or Erlang-Exponential model is straightforward:

```{r example, eval=FALSE}
library(GenErlangFit)
library(ggplot2)

# Default: Automatically fits both Erlang and Erlang-Exponential models,
# performs goodness-of-fit, and outputs the simplest statistically-acceptable model
result <- GenErlang_Fit(empiricaldata)
print(result)

# Fit only the Erlang model and performs goodness-of-fit
result_erlang <- GenErlang_Fit('Erlang', empiricaldata)
print(result_erlang)

# Fit only the Erlang-Exponential model and performs goodness-of-fit 
# (This option requires specifying an initial guess for K)
result_erlangexp <- GenErlang_Fit('ErlangExp', empiricaldata, K = 3)
print(result_erlangexp)
```
